Online learning methods for financial time series forecasting

### 2017-11-19 slides/defense-slides.tex
Motivation nearly finished

\begin{frame}[t,fragile]{Random walk and Cointegration Illustration}
 \begin{itemize}
  \item<1-> The drunkard's walk
  \item<2-> The dog's walk
  %\item<3-> The drunk and dog's walk
 \end{itemize}

@@

The relationship between non-stationary time series can be modelled if a stationary linear combination is shown to exist.
%Financial time series are mainly non-stationary.  However, one or more linear
%combinations of non-stationary time series can be stationary even though
%individually they are not. 
When this happens it is said they have a long-run
equilibrium relationship and the variables are {\color{red} cointegrated}.
\

@@

\Large
\begin{equation*}
 \Delta \mathbf{y}_t = \underbrace{
        \tikz[baseline]{
            \node[fill=red!20,ellipse,anchor=base] (t1)
            {$ \Omega\mathbf{y}_{t-1}$};
        }}_\text{Error correction term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=blue!20,anchor=base] (t2)
            {$\sum_{i=1}^{p-1} \phi_i^* \Delta \mathbf{y}_{t-i} $};
        }}_\text{Autorregresive term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=green!20,anchor=base] (t3)
            {$c$};
        }}_\text{Constant}
        +
        \underbrace{
        \tikz[baseline]{
            \node[fill=yellow!20,anchor=base] (t3)
            {$\mathbf{\epsilon}_t $};
        }}_\text{i.i.e}
\end{equation*}

@@

\begin{frame}
The Vector Error correction model (VECM) is used to model cointegrated time series but is used only with batch data and rarely used with high frequency data.
\end{frame}

\subsection{Proposal}
\begin{frame}
\frametitle{Proposal}
\begin{block}{}
To use VECM with high frequency data where the best parameters are updated when new data is available and do this is done in a fast way.
\end{block}
\end{frame}

%\section{Background}
%%
%%\begin{frame}
%%\frametitle{Integration }
%%\begin{block}{Integration of order $d$}
%%A time series $\mathbf{y}$ is said to be integrated of order $d$ if after
%%differentiating the variable $d$ times, we get an stationary process, more precisely:
%%\[
%%(1-L)^d \mathbf{y} \sim \text{I(0)} \, ,
%%\]
%%\noindent where I(0) is a stationary time series and $L$ is the lag operator:
%%\[
%%(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
%%\]
%%\end{block}
%%\end{frame}
%%
%%\subsection{Integration example}
%%\begin{frame}
%%\frametitle{Financial Time Series}
%%\begin{columns}
%%\column[t]{5cm}
%%High frequency financial time series are commonly an I(1) process 
%%\begin{center}
%%\includegraphics[width=0.85\textwidth]{img/EURUSD}
%%\end{center}
%%\column[t]{5cm}
%%Since their differences are stationary (I(0) process).
%%\begin{center}
%%\includegraphics[width=0.85\textwidth]{img/DEURUSD}
%%\end{center}
%%\end{columns}
%%\end{frame}
%
%\subsection{Cointegration}
%\begin{frame}
%\frametitle{Cointegration}
%\begin{block}{Cointegration definition}
%Let $\mathbf{y} = \{\mathbf{y}^1, \dots, \mathbf{y}^l\}$ be a set of $l$
%time series of order I(1) which are said to be cointegrated if a vector
%$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^l$  exists such that the
%time series,
%
%\begin{equation*}
% \mathbf{Z}_t:= \beta^\top \mathbf{y} = \beta(1) \mathbf{y}^1 + \dots + \beta(l) \mathbf{y}^l \sim
% \text{I(0)}\, .
%\end{equation*}
%\end{block}
%In other words, a set of I(1) variables is said to be cointegrated if
%a linear combination of them exists which is I(0).
%
%\end{frame}
%
%\subsection{Vector Autorregresive models}
%\begin{frame}
%\frametitle{VECM: Vector Error Correction model}
%
%VECM is a special form of a Vector Autorregresive model (VAR) for I(1)
%variables $\mathbf{y}$ that are also cointegrated. Let 
%
%\begin{equation*}
%\label{eq:variables}
%\mathbf{y}_t = 
%\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t}
%\end{bmatrix}^\top \, \qquad t=p+1, \dots N. 
%\end{equation*}
%
%\begin{block}{VAR}
%VAR($p$) model describes the behaviour of a set of $l$
%endogenous variables as a linear combination of their last $p$ values:
%{\color{blue}
%\begin{equation*}
%\label{eq:var}
% \mathbf{y}_t = \phi_1 \mathbf{y}_{t-1}  + \dots +   \phi_p\mathbf{y}_{t-p}
%+ \mathbf{c} + \mathbf{\epsilon}_t \, ,
%\end{equation*}}
%\noindent where  ${\phi_1,\dots,\phi_p}$ are $l \times l$
%matrices of real coefficients,
%$\mathbf{\epsilon}_{p+1},\dots,\mathbf{\epsilon}_N$ are error terms,
%$\mathbf{c}$ is a constant vector and $N$ is the total number of samples.
%\end{block}
%\end{frame}
%
%
%

>>>>

\begin{frame}
\frametitle{Stationarity}
\begin{block}{Strong definition}

The hole distribution of the variable does not depend on time, more precisely:
\small
\begin{equation*}
P\{y_{t_1} \leq
c_1,..,y_{t_L} \leq c_L\} = P\{y_{t_1+h} \leq c_1,..,y_{t_L+h} \leq c_L\}\enspace\forall L \in \mathbb{N}, \forall h \in \mathbb{Z}
\end{equation*}
\end{block}
\end{frame}

\begin{frame}
\frametitle{Stationarity}
\begin{block}{Weak definition}

A weakly stationary time series is a process which mean, variance and auto covariance do not change over time:
\small
{\color{blue}
 \begin{eqnarray*} E(y_t) &=& \mu  \quad
\forall t \in \mathbb{N} \\ E(y^2_t) &=& \sigma^2  \quad \forall t \in
\mathbb{N} \\ \lambda(s,t)&=&\lambda(s+h,t+h) \quad \forall s,t \in \mathbb{N},
\forall h \in \mathbb{Z} \end{eqnarray*}}
\end{block}
One of the consequences of stationary processes is how they recover from a shock. A shock represents an unexpected change in a variable or a in its error term in a particular period of time. For stationary time series, shocks to the system will gradually die away. 
\end{frame}

\begin{frame}
\frametitle{Non-stationary processes}
There are different types of non-stationary time series models often found in economics:
\begin{block}{Deterministic trend}
 Deterministic trend or trend stationary
processes have the following form:
{\color{blue}
\[
y_t = f(t) + \epsilon_{t}
\]}
\noindent where $\epsilon_{t}$ represents a stationary error term (with mean 0 and variance $\sigma^2$) and $f(t)$ is a deterministic function of time.
\end{block}
\end{frame}

\begin{frame}
\frametitle{Non-stationary processes}

\begin{block}{Stochastic trend}
Stochastic trend processes are also called {\color{red}unit root or I(1)} and have the following form:
{\color{blue}
\[
y_t = \mu + y_{t-1} + \epsilon_t
\]}
\noindent where $\epsilon_t$ is a stationary process. When $\mu = 0 $ the process is called pure random walk and when $\mu \neq 0$  the process is called random walk with drift.
\end{block}
\end{frame}

\begin{frame}
\frametitle{Integration }
\begin{block}{Integration of order $d$}
A time series $\mathbf{y}$ is said to be integrated of order $d$ if after
differentiating the variable $d$ times, we get an stationary process, more precisely:
\[
(1-L)^d \mathbf{y} \sim \text{I(0)} \, ,
\]
\noindent where I(0) is a stationary time series and $L$ is the lag operator:
\[
(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
\]
\end{block}
\end{frame}

\begin{frame}
\frametitle{Integration example}
\begin{columns}
\column[t]{5cm}
High frequency financial time series are commonly an I(1) process 
\begin{center}
\includegraphics[width=0.85\textwidth]{img/EURUSD}
\end{center}
\column[t]{5cm}
Since their differences are stationary (I(0) process).
\newline
\begin{center}
\includegraphics[width=0.85\textwidth]{img/DEURUSD}
\end{center}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Spurious regression}
Standard regression techniques are applied to non-stationary data.
The use of non-stationary data can lead to spurious regressions.

%\begin{block}{Spurious regression}
%If two stationary
%variables are generated as independent random series and are trending over time, when one of those variables is regressed on the other, they could have a high coefficient of determination ($R^2$) even if the two are totally unrelated. 
%\end{block}

\begin{columns}
\begin{column}{.45\linewidth}<1->
\begin{figure}
\includegraphics[width=0.4\paperwidth]{img/spurious1}
\end{figure}
\end{column}
\begin{column}{.45\linewidth}<2->
\begin{figure}
\includegraphics[width=0.4\paperwidth]{img/spurious2}
\end{figure}
\end{column}
\end{columns}
\end{frame}

\begin{frame}[t,fragile]{Random walk Illustration}
 \begin{itemize}
  \item<1-> The drunkard's walk
  \item<2-> The dog's walk
  %\item<3-> The drunk and dog's walk
 \end{itemize}

@@

The relationship between non-stationary time series can be modelled if a stationary linear combination is shown to exist.
%Financial time series are mainly non-stationary.  However, one or more linear
%combinations of non-stationary time series can be stationary even though
%individually they are not. 
When this happens it is said they have a long-run
equilibrium relationship and the variables are {\color{red} cointegrated}.
\end{frame}

\begin{frame}
\frametitle{Cointegration}
\begin{block}{Cointegration definition}
Let {\color{blue}$\mathbf{y} = \{\mathbf{y}^1, \dots, \mathbf{y}^l\}$} be a set of $l$
time series of order I(1) which are said to be cointegrated if a vector
$\beta=[\beta(1),\dots,\beta(l)]^\top \in \mathbb{R}^l$  exists such that the
time series,

\begin{equation*}
 \mathbf{Z}_t:= \beta^\top \mathbf{y} = \beta(1) \mathbf{y}^1 + \dots + \beta(l) \mathbf{y}^l \sim
 \text{I(0)}\, .
\end{equation*}
\end{block}
In other words, a set of I(1) variables is said to be cointegrated if
a linear combination of them exists which is I(0).

@@

\Large
\begin{equation*}
 \Delta \mathbf{y}_t = \underbrace{
        \tikz[baseline]{
            \node[fill=red!20,ellipse,anchor=base] (t1)
            {$ \Omega\mathbf{y}_{t-1}$};
        }}_\text{Error correction term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=blue!20,anchor=base] (t2)
            {$\sum_{i=1}^{p-1} \phi_i^* \Delta \mathbf{y}_{t-i} $};
        }}_\text{Autorregresive term} +
        \underbrace{
        \tikz[baseline]{
            \node[fill=green!20,anchor=base] (t3)
            {$c$};
        }}_\text{Intercept}
        +
        \underbrace{
        \tikz[baseline]{
            \node[fill=yellow!20,anchor=base] (t3)
            {$\mathbf{\epsilon}_t $};
        }}_\text{i.i.d}
\end{equation*}

@@

\begin{frame}
\begin{alertblock}{VECM limitation}
The Vector Error correction model (VECM) is used to model cointegrated time series but only with batch data and rarely used with high frequency data.
\end{alertblock}
\end{frame}

\subsection{Proposal}
\begin{frame}
\frametitle{Proposal}
\begin{block}{Proposal}
To use VECM with high frequency data where the best parameters are updated when new data is available and do this is done in a fast way.
\end{block}
\end{frame}

%\section{Background}
%
%
%\subsection{Cointegration}

%
%\subsection{Vector Autorregresive models}
%\begin{frame}
%\frametitle{VECM: Vector Error Correction model}
%
%VECM is a special form of a Vector Autorregresive model (VAR) for I(1)
%variables $\mathbf{y}$ that are also cointegrated. Let 
%
%\begin{equation*}
%\label{eq:variables}
%\mathbf{y}_t = 
%\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t}
%\end{bmatrix}^\top \, \qquad t=p+1, \dots N. 
%\end{equation*}
%
%\begin{block}{VAR}
%VAR($p$) model describes the behaviour of a set of $l$
%endogenous variables as a linear combination of their last $p$ values:
%{\color{blue}
%\begin{equation*}
%\label{eq:var}
% \mathbf{y}_t = \phi_1 \mathbf{y}_{t-1}  + \dots +   \phi_p\mathbf{y}_{t-p}
%+ \mathbf{c} + \mathbf{\epsilon}_t \, ,
%\end{equation*}}
%\noindent where  ${\phi_1,\dots,\phi_p}$ are $l \times l$
%matrices of real coefficients,
%$\mathbf{\epsilon}_{p+1},\dots,\mathbf{\epsilon}_N$ are error terms,
%$\mathbf{c}$ is a constant vector and $N$ is the total number of samples.
%\end{block}
%\end{frame}
%
%
%
